---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---

Introduction

This assignment is a component of the "Reproducible Research" course by Johns Hopkins University on Coursera. 

It is now possible to collect a large amount of data about personal movement using activity monitoring devices. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The raw data was provided as part of the course work. 
The repository is forked from [GitHub](http://github.com/rdpeng/RepData_PeerAssessment1).

## Loading and preprocessing the data

```{r, echo = FALSE, results= "hide"}

library(tidyverse)
library(lubridate)

```

```{r, echo = TRUE}

unzip(zipfile = "activity.zip")
activity <- read.csv2("activity.csv", sep = ",", na.strings = "NA")
activity <- activity %>% mutate(date = ymd(as.character(date)))

```

## What is mean total number of steps taken per day?

```{r, echo = TRUE}

# create tibble with total number of steps per day
totalsteps <- activity %>% 
        group_by(date) %>% 
        summarise(total = sum(steps))
```



```{r, echo = TRUE}

# calculate summary statistics
summary(totalsteps$total)

```

In the observation period, the monitored person took approximately 10,766 steps per day on average. The median number of steps taken is almost identical with 10,765. The daily number of steps taken varied widely, with a minimum of 41 and a maximum of 21,194 steps.

The distribution of steps is shown in here.

```{r, echo = TRUE}

# plot a histogram of total number of steps taken each day
ggplot(totalsteps,aes(total)) +
        geom_bar(na.rm = TRUE) + 
        scale_x_binned() + 
        labs(x = "Daily steps", y = "Number of days", 
             title = "Distribution of total steps taken per day",
             subtitle = "totaled over 2 months, 8 days without data")

```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
