---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---

Introduction

This assignment is a component of the "Reproducible Research" course by Johns Hopkins University on Coursera. 

It is now possible to collect a large amount of data about personal movement using activity monitoring devices. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The raw data was provided as part of the course work. 
The repository is forked from [GitHub](http://github.com/rdpeng/RepData_PeerAssessment1).

## Loading and preprocessing the data

```{r, echo = FALSE, results= "hide"}

library(tidyverse)
library(lubridate)

```

```{r, echo = TRUE}

unzip(zipfile = "activity.zip")
activity <- read.csv2("activity.csv", sep = ",", na.strings = "NA")
activity <- activity %>% mutate(date = ymd(as.character(date)))

```

```{r, echo = TRUE}

# create tibble with total number of steps per day
totalsteps <- activity %>% 
        group_by(date) %>% 
        summarise(total = sum(steps))
```


## What is mean total number of steps taken per day?

At first, we take a look at the activities.

Over the 2 month observation period, a total number of 570,608 steps were recorded. 

The daily activity was not spread evenly across the period. The daily average was slightly above 10,766 steps. The median number of steps taken was almost identical with 10,765. However, the daily number of steps taken varied widely, with a minimum of only 41 and a maximum of 21,194 steps.

```{r, echo = TRUE}

# calculate summary statistics
sum(totalsteps$total, na.rm = TRUE)
summary(totalsteps$total)

```


The below chart shows the distribution of daily activity.

```{r, echo = TRUE}

# plot a histogram of total number of steps taken each day
ggplot(totalsteps,aes(total)) +
        geom_bar(na.rm = TRUE) + 
        scale_x_binned() + 
        labs(x = "Daily steps", y = "Number of days", 
             title = "Distribution of total steps taken per day",
             subtitle = "totaled over 2 months, 8 days without data")

```


## What is the average daily activity pattern?

In a next step we take a look at the activity pattern over the course of an average day.
As can be seen, the activities start shortly after 5:00am, peak around 8:30am, start to calm down around 7:00pm, only to get close to zero around 10:00 pm.

```{r, echo = TRUE}

# create tibble with mean number of steps per interval
intraday <- activity %>%
  group_by(interval) %>%
  summarise(intra = mean(steps, na.rm = TRUE))

# find most active period
maxsteps <- intraday %>%
        filter(intra == max(intra))
maxsteps$interval

# plot a histogram of total number of steps taken each day
ggplot(intraday,aes(x = interval, y = intra)) +
        geom_line(na.rm = TRUE) + 
        labs(x = "Time (HHMM)", y = "Steps taken in interval", 
             title = "Average activity patter overe a day",
             subtitle = "averaged over 2 months, 5 minute intervals")

```

The most active interval is `r maxsteps$interval`.

```{r, echo = TRUE}

# find most active period
maxsteps <- intraday %>%
        filter(intra == max(intra))
maxsteps$interval

```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
