---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---

Introduction





```{r setoptions, echo = FALSE}

library(knitr)
opts_chunk$set(echo = TRUE, results = "asis")
options(scipen=999, digits = 2)

```

```{r libraries, echo = FALSE}

library(tidyverse)
library(lubridate)

```

This assignment is a component of the "Reproducible Research" course by Johns Hopkins University on Coursera. 

It is now possible to collect a large amount of data about personal movement using activity monitoring devices. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The raw data was provided as part of the course work. 
The repository is forked from [GitHub](http://github.com/rdpeng/RepData_PeerAssessment1).

## Loading and preprocessing the data

```{r load_data}

unzip(zipfile = "activity.zip")
activity <- read.csv2("activity.csv", sep = ",", na.strings = "NA")
activity <- activity %>% mutate(date = ymd(as.character(date)))

```

```{r preprocess}

# create tibble with total number of steps per day
totalsteps <- activity %>% 
        group_by(date) %>% 
        summarise(total = sum(steps, na.rm = TRUE))

totalsum <-  sum(totalsteps$total, na.rm = TRUE)
totalmean <-  mean(totalsteps$total, na.rm = TRUE)
totalmedian <- median(totalsteps$total, na.rm = TRUE)
totalmin <- min(totalsteps$total, na.rm = FALSE)
totalmax <- max(totalsteps$total, na.rm = TRUE)

```


## What is mean total number of steps taken per day?

At first, we take a look at the activities.

Over the 2 month observation period, a total number of `r totalsum` steps were recorded. 

The daily activity was not spread evenly across the period. The daily average was `r totalmean` steps. The median number of steps taken was `r totalmedian`. However, the daily number of steps taken varied widely, with a minimum of `r totalmin` and a maximum of `r totalmax` steps.

The below chart shows the distribution of daily activity. 

```{r histo_total}

# plot a histogram of total number of steps taken each day
ggplot(totalsteps,aes(total)) +
        geom_bar(na.rm = TRUE) + 
        scale_x_binned() + 
        labs(x = "Daily steps", y = "Number of days", 
             title = "Distribution of total steps taken per day",
             subtitle = "totaled over 2 months, 8 days without data")

# Raw data NAs show up in the first bin, as set to 0.
# Would have been better to remove them from the sample before creating the plot. However, the instructions stated these could be ignored.

```



## What is the average daily activity pattern?

In a next step we take a look at the activity pattern over the course of an average day.
As can be seen, the activities start shortly after 5:00am, peak around 8:30am, start to calm down around 7:00pm, only to get close to zero around 10:00 pm.

```{r histo_mean}

# create tibble with mean number of steps per interval
intraday <- activity %>%
  group_by(interval) %>%
  summarise(intra = mean(steps, na.rm = TRUE))

# plot a histogram of total number of steps taken each day
ggplot(intraday,aes(x = interval, y = intra)) +
        geom_line(na.rm = TRUE) + 
        labs(x = "Time (HHMM)", y = "Steps taken in interval", 
             title = "Average activity patter overe a day",
             subtitle = "averaged over 2 months, 5 minute intervals")

```

```{r max_steps}

# find most active period
maxsteps <- intraday %>%
  filter(intra == max(intra)) %>%
  mutate(stunde = interval %/% 100, minute = interval %% 100)

```
The most active 5-minute-interval starts `r maxsteps$stunde` : `r maxsteps$minute`.


## Imputing missing values

In a next step we consider missing values and their potential effects  on the results achieved so far. 

```{r missing}

totalnum <- sum(is.na(activity$steps))

# create tibble to analyze missing values
miss <- activity %>%
  filter(is.na(steps)) 

miss_days <- miss %>%
  group_by(date) %>% 
  count(date) 
# 8 days with 288 intervals each = 2304 observations are completely missing.

```

There are in total `r totalnum` intervals with missing observations. Those missing values fall  on only eight days. None of these days has any measurement whatsoever.

It might be required to impute missing values.
Given observed intraday activity pattern it seems prudent to replace missing values with the mean for respective 5-minute intervals.

```{r impute}

# create new dataset where NAs are replaced with interval mean 
imputed <- activity %>% 
        group_by(interval) %>%
        mutate(steps = replace(steps, is.na(steps), mean(steps, na.rm = TRUE)))

# From here on missing values initially set to 0 have been replaced 
# create tibble with total number of steps per day
totalimputed <- imputed %>% 
        group_by(date) %>% 
        summarise(total = sum(steps))

imputedmean <- mean(totalimputed$total) 
imputedmedian <-  median(totalimputed$total) 


# plot a histogram of total number of steps taken each day
ggplot(totalimputed,aes(total)) +
        geom_bar() + 
        scale_x_binned() + 
        labs(x = "Daily steps", y = "Number of days", 
             title = "Distribution of total steps taken per day",
             subtitle = "totaled over 2 months, missing values imputed")
```

As one would expect, replacing missing values with the mean of the non-missing values moves observations to center of the distribution. This may lead to different results. 

The new mean number of steps with imputed data is `r imputedmean`. The new 
median number of steps with imputed data is `r imputedmedian`. 


## Are there differences in activity patterns between weekdays and weekends?
